% This template is public domain.
\documentclass{ltugboat}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{ifpdf}
\ifpdf
\usepackage[breaklinks,hidelinks]{hyperref}
\else
\usepackage{url}
\fi

% non-ASCII characters needed
\usepackage[utf8]{inputenc}

% Bibliography contains Slovak language
\usepackage[slovak]{babel}

% Bibliography setup.
\usepackage[style=iso-numeric]{biblatex}
\addbibresource{../resource.bib}

% Define command for BibLaTeX
\newcommand\BibLaTeX{\textsc{Bib}\LaTeX}

%%% Start of metadata %%%

\title{Typesetting Bibliographies Compliant with the International
	Standard ISO~690 in \LaTeX{}}

% repeat info for each author.
\author{Dávid Lupták}
\address{Faculty of Informatics \\ Masaryk University \\ Brno, Czech Republic}
\netaddress{422640@mail.muni.cz}
% \personalURL{http://example.org/~user/}

%%% End of metadata %%%

\begin{document}

\maketitle

\begin{abstract}

The preparation of bibliographic references and citations compliant
with the international standard ISO~690 is required by many institutes
not limited to the Czech and Slovak academia. However, the typesetting
of bibliographies conforming to the respective standard is not yet
supported in the \LaTeX{} document preparation system. The
\verb|biblatex-iso690| package has been revised and improved to fully
meet the requirements of the international standard and thus greatly
simplifies the typesetting of bibliographies for all kinds of
information resources.

\end{abstract}

\section{Introduction}

Writing an article or paper or any other kind of work requires high
maturity to work with other resources which should be also referenced
and cited properly. The preparation of bibliographic references and
citations is mainly required to comply with the international standard
ISO~690 in Czech academia~\cite{Kratochvil2011}. At the beginning
of this article the standard ISO~690 is introduced briefly, followed
by various existing software implementions incorporating the standard.
Typesetting of bibliographies in \LaTeX{} preparation system is next
covered in more detail and the package \textsf{biblatex-iso690},
the first complete implementation in \LaTeX{} which is compliant with
the latest version of the standard ISO~690, is featured.


\section{International standard ISO~690}

The preparation of bibliographic references and citations
was done in accord with the international standard
ISO~690:1987~\cite{iso690:1987} for printed resources and
ISO~690-2:1997~\cite{iso690:2:1997} for electronic information
resources in the past. These two versions of the standard were
unified and replaced by new version ISO~690:2010~\cite{iso690:2010}
in 2010. On the national level, translations of such standards are
provided by offices for standards
(ISO members)~\cite{isoMembershipManual} with the equal status to
the original standard. Examples of such translations are
Czech ČSN ISO~690:2011~\cite{csn:iso690:2011},
Slovak STN ISO~690:2012~\cite{stn:iso690:2012} or
German DIN ISO~690:2013~\cite{din:iso690:2013}.


\subsection{Terminology}

Let's briefly introduce two fundamental terms and definitions of the
standard~\cite{iso690:2010} which are essential for understanding
following text properly:

\begin{description}
\item[citation] indication within the text or other form of content
      of a relevant reference
\item[reference] data describing a resource or part thereof,
      sufficiently precise and detailed to identify it and to enable
      it to be located
\end{description}



\subsection{Consistency principle}

The international standard ISO~690 does not prescribe a particular
style of reference or citation. The examples used in the standard
are not prescriptive as to style and punctuation. These facts outline
two findings:

\begin{enumerate}
\item separation of semantic and presentation is respected % preserved
\item the standard cannot be considered as a citation style~\cite{Hala2013}
\end{enumerate}

\noindent At the same time, a uniform style, format and punctuation
scheme shall be used for all references in a document, regardless of
the particular style guide being used. Fulfillment of this requirement
is left to creator of references, with examples illustrated in the
standard itself, various national interpretations ar just typical
usages of typesetting a bibliography.


\section{Typesetting of bibliographies in \LaTeX{}}

\LaTeX{} document preparation system provides three possibilities of
typesetting bibliographies~\cite{talbot2013}. The first approach is to
use just the plain \LaTeX{} to generate the bibliographies, two others
stick to the separation of semantic and presentation principle and
take benefit of creating an external database of bibliographic data
and using an application to generate the output.


\subsection{Plain \LaTeX{}}

The \verb|thebibliography| environment for references and the
\verb|\cite| command family for citations are available in \LaTeX{}.
Each single reference is then mentioned as \verb|\bibitem| with its
unique identifier in the \verb|thebibliography| environment.

\begin{verbatim}[\small]
\documentclass{...}
\begin{document}
\cite{<label01>}
...
\begin{thebibliography}{<widest label>}
\bibitem{<label01>}
  <Author>. \emph{<Title>: <subtitle>}. ...
...
\end{thebibliography}
\end{document}
\end{verbatim}

The aforementioned code snippet shows the basic syntax of this
approach and also unveil the fact that it is not practical with a
large number of citations~\cite{talbot2012}. The main drawbacks are:

\begin{enumerate}
\item all bibliography entries listed in \verb|thebibliography|
  environment are typeset (regardless of whether they are cited
  or not)\label{item:1}
\item every bibliography entry has to be entered and formatted
  manually for every desired bibliography style\label{item:2}
\item bibliography references are not sorted at all (they are printed
  out in the order as they are listed in \verb|thebibliography|
  environment)\label{item:3}
\end{enumerate}

\noindent Standard ISO~690 does not prescribe any guidelines to deal
with the disadvantage number~\ref{item:1}, however such results do not
follow general recommendations for creating
bibliographies~\cite{talbot2013}. Regarding the
limitation~\ref{item:2} it is really tough task to ensure the
consistency of references and regarding the drawback~\ref{item:3} it
is not possible at all to output references in the correct order for
any method of citation introduced in the standard.

Reusability of bibliographic entries and scalability of the list
of references lack in this approach. On the other hand, one of the
main strenghts is relatively fast document compilation (it is enough
to compile it twice by \TeX{} engine).


\subsection{B\textsc{ib}\TeX{}}

The preferred method of generating a bibliography is to create an
external bibiliography database (see section~\ref{bib:file}) and use
an application to generate the output~\cite{talbot2013}. Such
applications can deal with typesetting references in the correct
order (solves the issue~\ref{item:3}) and based on the selected
bibliography style (solves the issue~\ref{item:2}) they generate
\verb|thebibliography| environment which then can be input into the
document. One typical representative of this method is \BibTeX{},
which adhere to the principle of separation of semantic and
presentation.

The \verb|\bibliographystyle| command is used to define the desired
bibliography formatting style, the \verb|\bibliography| command
specifies external bibliography database to use and also the location
where the list of references is to be printed. The \verb|\cite|
family commands are used to create a citation within a text of a
document pointing to the reference in question. It is also possible
to use \verb|\nocite| command, which adds the bibliography entry to
the list of references without printing a citation within a text
(addressing problem~\ref{item:1}).

\begin{verbatim}[\small]
\documentclass{...}
\bibliographystyle{<formatting style>}
\begin{document}
\cite[<additional info>]{<list of labels>}
...
\bibliography{<list of database files>}
\end{document}
\end{verbatim}

This brief introduction to \BibTeX{} seems to be promising, however
there is also a lot of troubles out there underlined by the fact,
that development of the \BibTeX{} package is
stagnant~\cite{patashnik1994,patashnik2003}. Let's list the main
disadvantages and limitations:

\begin{enumerate}
\item input encoding problems~\cite{ctan:bibtex} (even though an
  alternative solution is
  available)\footnote{\url{https://www.ctan.org/pkg/bibtex8bit}}
\item designing your own \BibTeX{} styles is rather
  difficult~\cite{patashnik1988:styles} (even though an alternative
  solution for generating \BibTeX{} styles is
  available)\footnote{\url{https://www.ctan.org/pkg/custom-bib}}
\item capacity issues (working with large bibliography database files,
  BibTeX is very likely to run out of memory)~\cite{biblatex:manual2016}
\item lacking of citation customizations~\cite{bibtex:faq} (even
  though more flexible solutions are
  available)\footnote{\url{https://www.ctan.org/pkg/natbib},
  \url{https://www.ctan.org/pkg/cite}}
\item missing of contemporary fields widely used nowadays, e.g.
  the \verb|url| field (even though an alternative solution is
  available)\footnote{\url{https://www.ctan.org/pkg/natbib},
  \url{https://www.ctan.org/pkg/babelbib}}
\item lacking of translations and multilingual
  bibliographies~\cite{Harders2002} (even though an alternative
  solution is
  available)\footnote{\url{https://www.ctan.org/pkg/babelbib}}
\end{enumerate}

To typeset your document properly, it is needed to compile your
document at least three times by \TeX{} engine and at least once
by \BibTeX{} program. Here is the global pattern to be
applied~\cite{Markey2009}:

\begin{center}
\LaTeX\ (\BibTeX\ \LaTeX)$^+$ \LaTeX\
\end{center}

\noindent Generating a bibliography using \BibTeX{} in comparison
with the plain \LaTeX{} introduces more complexity, however, on the
other hand, it also deals with almost all of the aforementioned
limitations succesfully.


\subsection{\BibLaTeX}\label{biblatex:typesetting}

Another option for generating a bibliography using external
bibliography database and application for compiling it reveals in
the \BibLaTeX{} package of \LaTeX{}. This package is a complete
reimplementation of the bibliographic facilities provided by \LaTeX{},
usually referred to as a successor of an ancient \BibTeX{}
package~\cite{ctan:bibtex,hufflen2011}. Formatting of the
bibliography is entirely controlled by \TeX{} macros and
processing a bibliography database file (see also~\ref{bib:file})
the \verb|biber| backend program is used~\cite{biblatex:manual2016}.

The usage of \BibLaTeX{} package slightly differs from traditional
\BibTeX{} since it provides also more advanced bibliographic
facilities for use with \LaTeX{}. From the user perspective,
different syntax is noticable. Formatting styles are specified as
a load-time package option in the optional argument to
\verb|\usepackage|. Bibiliography database files are specified in
the document preamble with \verb|\addbibresource| command using the
full name of the file (including \verb|.bib| extension). The list
of references is printed using the \verb|\printbibliography| command
at the position where this command is placed in a document.
For creating citations within a text of a document, \verb|\cite|
command and its variants are used. Basic structure is as follows:

\begin{verbatim}[\small]
\documentclass{...}
\usepackage[...]{biblatex}
\addbibresource{<database01.bib>}
\addbibresource{<database02.bib>}
\begin{document}
\cite{...}
...
\printbibliography
\end{document}
\end{verbatim}

\BibLaTeX{} successfully deals with a lot of limitations introduced
by \BibTeX{}, to mention a few of them~\cite{ctan:biblatex}:

\begin{enumerate}
\item full Uni­code sup­port
\item highly cus­tomis­able sort­ing and bibliography labels
\item \verb|polyglossia| and \verb|babel| supp­port for au­to­matic
  lan­guage switch­ing for bib­li­o­graphic en­tries and ci­ta­tions
\item more entry types and fields
\item a lot of bibliography and citation styles already available
\item easy to de­sign new bib­li­og­ra­phy and ci­ta­tion styles
\end{enumerate}

The list could continue to be extended to the rich functionality
provided by the \BibLaTeX{} package~\cite{biblatex:manual2016}. On
the other hand, there are almost none drawbacks of using the package.
To mention just one for all \dash incompatibility of \BibLaTeX{}
auxiliary files when submitting to a
journal~\cite{stack:ex:biblatex:drawback}.

Document compiling is analogous to \BibTeX{} approach. At first,
document is compiled by \TeX{} engine followed by running \verb|biber|
on a generated auxiliary \verb|.bcf| file and then compile by \TeX{}
engine once again. The \BibLaTeX{} schema for compiling a document
is as follows (note that file extensions are optional):

\begin{verbatim}[\small]
latex <document>[.tex]
biber <document>[.bcf]
latex <document>[.tex]
\end{verbatim}


\subsection{Bibliography database (\texttt{.bib} file)}\label{bib:file}

For the sake of completness of this section it is desireable to
introduce also the bibliography database \verb|.bib| file. This file
contains bibliography entries, each entry is of specific type (the
word after @), has unique label and a number of tags (key-value
pairs) defining resource data. General schema of an entry looks
like the following~\cite{bib:format}:

\begin{verbatim}[\small]
@<entry type>{<label>,
  <field> = {<value>},
  ...
  <field> = {<value>}
}
\end{verbatim}

\noindent 

All of the entry types supported by \BibTeX{} can be used directly
or via alias also with \BibLaTeX{} package. On top of traditional
types, \BibLaTeX{} package introduces other ones with the possibility
to define completely new ones.

The same situation applies also for entry fields. \BibLaTeX{} package
provides backward compatibility with all of the \BibTeX{} fields and
on top of them it adds some more. In addition to regular fields there
are available so-called special fields which can contain additional
settings related to an entry, e.g. to specify the language on per
entry basis for mulitilingual bibliographies.


\subsection{Summary}

Basic functionality of \LaTeX{} for generating a bibliography can be
appropriately used for small number of citations in a document.
However, in case of a large number of citations, it is best to use
an external bibliographic application. This approach stick to a
principle of separation semantic and presentation, which results in
high scalability and reusability of bibliography entries and
makes the working with references more flexible and efficient.

Besides \BibTeX{} \dash traditional representative of this method
\dash many other applications based on it exist out there. However,
all of them are limited to the fact that they are based on \BibTeX{}.
Mainly it is a case of used formatting styles, even though some of the
applications try to replace the language BST (\BibTeX{} STyle) and use
more modern languages \dash mostly
\XML{}~\cite{hufflen2011,hufflen2008}.

As it turned out from the big variety of options for typesetting
a bibliography in \LaTeX{}~\cite{talbot2013,Mittelbach2004}, best
choice nowadays is to use \BibLaTeX{} package with its backend
application \verb|biber|~\cite{hufflen2011,biber:manual2016}.


\section{ISO~690 implementations}

This section introduces an existing software products, tools or
services which incorporate the standard ISO~690. First two
mentioned here are designed to use with \LaTeX{} document preparation
system, language CSL is covered thanks to its gaining usage
popularity and \verb|OPmac-bib| package because it is a rare example
among all available packages which deliver the full support for this
particular standard. More comprehensive overview of the available
solutions can be found in the bachelor thesis of the author of this
article in Slovak~\cite{Luptak2016}.


\subsection{czechiso}

For the Czech versions of the standard \dash
ČSN ISO 690:1996~\cite{csn:iso690:1996} and
ČSN ISO~690-2:2000~\cite{csn:iso690:2:2000} \dash is  available an
unofficial formatting style \verb|czechiso| created by David Martinek
in 2006. The style can be found
at \url{http://www.fit.vutbr.cz/~martinek/latex/czechiso.html}. This
implementation does not meet the requirements of the standard
precisely, it is lacking some required fields for bibliographic
entries and many functions responsible for printing out a reference
correctly are in need of rewriting to fully conform to the standard.


\subsection{biblatex-iso690}

First version of the bibliography and citation style for \BibLaTeX{}
conforming to the standard ISO~690 is dated back to 2011. This
implementation was based on the previous versions of the
standards~\cite{csn:iso690:1996,csn:iso690:2:2000} and on the Czech
interpretation~\cite{Bratkova2008}. The package was created by Michal
Hoftich and it is available
at \url{https://github.com/michal-h21/biblatex-iso690}. Similarly to
the previous aforementioned package, even this solution did not
adhere to the standards precisely. Many issues related to the
functionality of the package as well as the usage of this style were
reported on the homepage of the project. Thanks to reviewing the
style completely in 2016, the package fully meet the requirements
of the standard at the current stage (see also
subsection~\ref{biblatex-iso690}).


\subsection{Language CSL}

The Citation Style Language (CSL) is an open \XML{} based language for
working with bibliographies. It became popular with the release of
Zotero reference manager in 2006~\cite{Fenner2010}.

The main advantage of this language is the syntax of \XML{}, jointly
with the popularity, open source initiative and versatility of
CSL~\cite{Ansorge2013}. Another undisputed benefit is its widespread
character, which is indicated by extensive list of products which use
CSL styles on the official webpage of the CSL
project~\cite{csl:home}. One of the most famous are Zotero, Papers
and Mendeley.

In the CSL style repository can be found over 8000 styles, among
which is available also 15 styles for the standard ISO~690. These
styles differs in localisation or method of citation, hence so huge
number of styles just for one standard. In all of them appear minor
errors or inaccuracies in accord with the standard ISO~690. In
general, CSL has a number of limitations~\cite{csl:styles}:

\begin{itemize}
\item limited support for customizing the label format
\item limited support for legal styles (Multilingual Zotero
  as an alternative)
\item limited support for citing items in multiple languages within
  a single document (Multilingual Zotero as an alternative)
\item limited support for entering date range into the date field
  (entry is not generated)
\end{itemize}

In conclusion, it should be added that \BibLaTeX{} package covers
all of these limitations in its basic
functionality~\cite{biblatex:manual2016}.


\subsection{OPmac-bib}

OPmac are additional macros on top of plain \TeX{} taking advantage
of basic \LaTeX{} functionality. The additional package OPmac-bib is
built as a part of it, available for bibliography facilities. No
external program for generating a bibliography is needed, all is
handled by \TeX{} macros and the \verb|librarian.tex| package by
Paul Isambert. The OPmac package is created by Petr Olšák and it
ships with \verb|csplain| package since 2015. More details about
OPmac package are left to another article of \TUB{}~\cite{olsak2016}.

OPmac-bib can process all of the traditional types and fields of
\BibTeX{} and above and beyond it introduces new fields, which are
a must when working with bibliographies nowadays. These fields are
e.g. \verb|url|, \verb|doi| or \verb|lang|, which bypass the need
of using \verb|note| field for providing such data in this field.
Hence it is possible to print out this data in the correct order in
accord with the standard.

While \BibTeX{} was lacking a lot of required types and fields,
OPmac-bib is making much better. But still the standard is so complex
that even OPmac-bib does not cover all of the pitfalls the standard
introduces regarding the fields. However, OPmac-bib can deal with
this really smartly. The package provides a few versatile fields
which can be used for entering bibliographic data jointly with the
formatting macros to customize the field. Hence one can achieve
desired output. E.g. \verb|option| and \verb|ednote| are examples of
such fields.

The field \verb|option| can be used for entering other titles,
translation of title etc. Thanks to this field the correct output
conforming to the latest version of the standard can be achieved.

The field \verb|ednote| can be used for entering secondary authors
or other additional information. Note that formatting of this field
is not further processed, so the entered value is printed out as is.
Hence one has to be careful to conform to the standard when entering
the data. Typical examples of entered data are translators or
originators of multiple editions.

Thanks to the availability of these non-standard fields and full
customization is the OPmac-bib package an obvious and preferred
choice to work with plain \TeX{}.


\section{Package biblatex-iso690}\label{biblatex-iso690}

Among all implementations incorporating the standard ISO~690
mentioned in the previous section, only one is relevant to work on
for typesetting a bibliography in \LaTeX{} \dash the
\verb|biblatex-iso690| package. Original implemetation deviated from
the standard, but since its review, \verb|biblatex-iso690| package
is fully compliant with the latest version of the international
standard ISO~690.

Original state of the \verb|biblatex-iso690| package contained the
following defects and drawbacks:

\begin{itemize}
\item adhered to out-of-date editions of the standard
\item incorrect order of elements
\item (redundant/missing) punctuation
\item missing some types of resources
\item missing some required elements
\item missing creator secondary responsibility
\item obsolete and deprecated code
\end{itemize}

Initial analysis of the original state of the \verb|biblatex-iso690|
package resulted in its complete reimplementation. Correction of
printing bibliography elements in the correct order in a
reference was not the only change. Almost all macros, commands and
definitions for parsing fields from the \verb|.bib| database file
needed to be refactored. A lot of requirements of the standard could
be achieved simply by the author interface of the \BibLaTeX{}
package. For some requirements was needed to refine a few low-level
macros of the package. And some were left just to an author of a
bibliography database, because they cannot be solved algorithmically.
Known limitations are:

\begin{itemize}
\item missing support for running notes citation method
\item URL addresses wrapping
\item algorithmic solution for (not) printing out a first edition of
  a resource
\item algorithmic solution for (not) printing out only first
  publisher
\item algorithmic solution for (not) printing out only first
  places
\item term \textit{Anon} for anonymous works
\item localisation string \textit{nodate} for no date
\end{itemize}


\subsection{Methods of citation}

The standard ISO~690 prescribes three citation methods of information
resources. Besides aforementioned running notes method it is also
so-called Harvard system (known also as author-date) and numeric
system. In the \verb|biblatex-iso690| package they are available as
\verb|iso-authoryear| and \verb|iso-numeric|. Note that formatting
style is specified as a package option of \BibLaTeX{} package, e.g.\\
\verb|\usepackage[style=iso-numeric]{biblatex}|.


\subsection{Package options \dash customization}

The standard ISO~690 does not prescribe any particular style, format
and punctuation scheme for references to be used. Frequently demanded
changes are available as \verb|biblatex-iso690| package options,
which are:

\begin{itemize}
  \item \verb$spacecolon=[true|false]$ to change printing of colon in
    subtitles and publication information
    \begin{itemize}
    \item Place : Publisher
    \item Place: Publisher
    \end{itemize}
  \item \verb$pagetotal=[true|false]$ to print out a total number of
    pages of a resource as an optional information
    \begin{itemize}
    \item Place : Publisher, 2008 [60 p.]
    \item Place : Publisher, 2008
    \end{itemize}
  \item \verb$shortnumeration=[true|false]$ to distinguish numeration
    and pagination typographically
    \begin{itemize}
    \item \dots\ 2011, \textbf{32}(3), 289\dash301 [visited on
    	2016-05-14] \dots
    \item \dots\ 2011, vol. 32, no. 3, pp. 289\dash301 [visited on
    	2016-05-14] \dots
    \end{itemize}
  \item \verb$thesisinfoinnotes=[true|false]$ to specify position of
    thesis information
    \begin{itemize}
    \item \dots\ Available from: \meta{url}. BT. MU, FI, Brno.
    	Supervisor Petr SOJKA
    \item \dots\ BT. MU, FI, Brno. Supervisor Petr SOJKA. Available
    	from: \meta{url}
    \end{itemize}
\end{itemize}


\subsection{Integration into fithesis3 class}

\verb|Fithesis3| is an official doc­u­ment class for the type­set­ting
of the­ses at the Masaryk Univer­sity (Brno, Czech Repub­lic) in
\LaTeX{}~\cite{novotny2015}. The class has been de­signed for easy
ex­ten­si­bil­ity by style and lo­cale files of other aca­demic
in­sti­tu­tions. It was also natural request to integrate the
\verb|biblatex-iso690| package into \verb|fithesis3| template. This
integration has been done by cooperation with the maintainer of the
\verb|fithesis3| package \dash Vít Novotný \dash and consists of the
following steps:

\begin{itemize}
\item the \verb|bib| key added to the package metadata section, which
  can be used to specify a list of \verb|.bib| database files
\item citation method is loaded automatically based on the selected
  faculty
\item list of references is printed automatically at the end of a
  document
\item all bibliography management can be also set up manually (see
  section~\ref{biblatex:typesetting})
\end{itemize}

\begin{verbatim}[\small]
\documentclass{fithesis3}
\thesissetup{
  ...
  bib = {<bibliography-database.bib>}
  ...
}
\begin{document}
\cite{...}
...
\end{document}
\end{verbatim}

\subsection{Availability}

As already mentioned, there was no official support for the standard
ISO~690 in \LaTeX{} so far. However,
\verb|biblatex-iso690| package has acquired an official status after
the revision and now is available from \CTAN{} as the package
\verb|biblatex-iso690|. Under the same name it is available also in
\TeX{} Live distribution since \TeX{} Live 2016.

\section{Summary}

The article describes typesetting a bibliography in \LaTeX{}
compliant with the international standard ISO~690. The standard is
introduced at the beginning followed by three possibilities of
typesetting a bibliography in \LaTeX{} being considered. Number of
implementations incorporating the standard ISO~690 exist but the
\verb|biblatex-iso690| package takes the most interest, because
since its initial implementation in 2011, the package was revised
in 2016 to fully meet the requirements of the most recent version of
the standard. References in this article are generated using the
reimplemented package \verb|biblatex-iso690| so they can serve for
the purpose of demonstration.


\printbibliography

\makesignature
\end{document}
